Estimating duration in the presence of misclassification

### 2024-02-19 ATACCC.tex
proof ch 4

The model of \textcite{hakkiOnset} has two components: a latent viral load model and an observation model.
The viral load model is a deterministic function of the model parameters, producing an individual's viral load trajectory.
The observation model mathematically formalises how the RT-PCR testing process produces noisy observations of the latent viral load.
The observation model includes the possibility of false positive and negatives, and the limit of detection of the RT-PCR test (see \cref{E-biology-data:sec:PCR}).
% The limit of detection means that viral loads below a certain threshold are not detected by the RT-PCR test.
% They can be represented as a censoring of the viral load at the limit of detection.

@@

The noisy observations of viral log load from RT-PCR testing are modelled by assuming each measurement of log viral load has an iid error distributed $\N(0, \sigma_v^2)$.
The measurements are left-censored at the limit of detection, $\vlod$, meaning that any observation which would give a log viral load of less than $\vlod$ is instead observed as negative.
False positives and negatives are modelled by allowing a test failure with probability $\rho$.
Test failures are modelled as producing a log viral load reading drawn from $\N(x_0, \sigma_0^2)$.
These readings are censored in the same way as true readings.
A censored test failure represents a false negative, otherwise the test failure represents a false positive.
\textcite{hakkiOnset} uses a prior for $x_0$ centred on $\vlod$, which encodes a prior belief that false negatives and false positives occur with equal probability.

@@

Individuals are split into groups based on their vaccination status.
The first group is unvaccinated and infected with a variety of pre-Alpha, Alpha, and Delta variants.
The second group is vaccinated and all were infected by the Delta variant.
My focus is on the first group since they best correspond to the infections in the time period used throughout this thesis.

@@

\textcite{hakkiOnset} uses weakly informative priors, similar to those previously used~\autocite{singanayagamCommunity}.
These are given in \cref{ATACCC:table:hakki-priors}.
I reran \textcite{hakkiOnset}'s analysis using R~4.3.1~\autocite{R-4-3-1}.
The inference was performed using RStan~2.32.3~\autocite{RStan-2-32-3} and Stan~2.32.2~\autocite{Stan-2-32-2}.
Inference settings were the same as \textcite{hakkiOnset}, which included 8 chains with a total of 8000 iterations each with the first 3000 used as warm-up.
\begin{table}
\begin{tabular}{l c l l}
    Parameter description & Symbol & Distribution \\
    \hline \\
    Mean max log viral load & $\mu_{v_\text{max}, g}$ & $\N(15, 15^2)$ \\
    Standard deviation of max log viral load & $\sigma_{v_\text{max}}$ & $\N_+(0, 10^2)$ \\
    Mean rate of log viral load increase & $\mu_{a, g}$ & $\N(1.25, 0.75^2)$ \\
    Standard deviation of rate of log viral load increase & $\sigma_{a}$ & $\N_+(0, 1)$ \\
    Mean rate of log viral load decrease & $\mu_{b, g}$ & $\N(0.5, 1.4^2)$ \\
    Standard deviation of rate of log viral load decrease & $\sigma_{b}$ & $\N_+(0, 1)$ \\
    Correlation of viral load parameters & $\matr{C}_\phi$ & $\LKJ(1)$ \\
    Negative log of rate of test failure & $-\log \rho$ & $\N(3, 0.5)$ \\
    Mean log viral load for test failures & $x_0$ & $\N(0, 3^2)$ \\
    Standard deviation of log viral load for test failures & $\sigma_0$ & $\N_+(3, 3^2)$ \\
    Standard deviation of log viral load observations & $\sigma_v$ & $1 + \N_+(2, 3^2)$ \\
    Time from first test to peak & $\tau_{\max,i}$ & $\N(5,5)$
\end{tabular}
\caption[Viral load model priors]{Priors for each parameter, taken from \textcite{hakkiOnset}. $\N_+$ indicates a normal distribution truncated to the non-negative reals. Details of the distributions, including their parameterisations, are in \cref{E-distributions}. The elements of $\vec{\mu}_{\phi,g}$ and $\vec{\sigma}_\phi$ have independent priors placed on them, as specified in the first lines of the table. Priors are the same for all $g$ and $i$. \label{ATACCC:table:hakki-priors}}
\end{table}

>>>>

The model of \textcite{hakkiOnset} has two components: a latent viral load model and an observation model.
The viral load model is a deterministic function of the model parameters, producing an individual's viral load trajectory.
The observation model mathematically formalises how the RT-PCR testing process produces noisy observations of the latent viral load.
The observation model includes the possibility of false positive and negatives, and the limit of detection of the RT-PCR test (\cref{E-biology-data:sec:PCR}).
% The limit of detection means that viral loads below a certain threshold are not detected by the RT-PCR test.
% They can be represented as a censoring of the viral load at the limit of detection.

@@

The noisy RT-PCR observations of viral log load are modelled by assuming each measurement of log viral load has an iid error distributed $\N(0, \sigma_v^2)$.
The measurements are left-censored at the limit of detection, $\vlod$, meaning that any observation which would give a log viral load of less than $\vlod$ is instead observed as negative.
False positives and negatives are modelled by allowing a test failure with probability $\rho$.
Test failures are modelled as producing a log viral load reading drawn from $\N(x_0, \sigma_0^2)$.
These readings are censored in the same way as true readings.
A censored test failure represents a false negative, otherwise the test failure represents a false positive.
\textcite{hakkiOnset} uses a prior for $x_0$ centred on $\vlod$, which encodes a prior belief that false negatives and false positives occur with equal probability.

@@

Individuals are split into groups based on their vaccination status.
The first group is unvaccinated and infected with a variety of pre-Alpha, Alpha, and Delta variants.
The second group is vaccinated and all were infected by the Delta variant.
My focus is on the first group since they most closely correspond to the infections in the time period used throughout this thesis.

@@

\textcite{hakkiOnset} uses weakly informative priors, similar to those previously used~\autocite{singanayagamCommunity}, see \cref{ATACCC:table:hakki-priors}.
I reran \textcite{hakkiOnset}'s analysis using R~4.3.1~\autocite{R-4-3-1}.
The inference was performed using RStan~2.32.3~\autocite{RStan-2-32-3} and Stan~2.32.2~\autocite{Stan-2-32-2}.
Inference settings were the same as \textcite{hakkiOnset}, which included 8 chains with a total of 8000 iterations each with the first 3000 used as warm-up.
\begin{table}
\begin{tabular}{l c l l}
    Parameter description & Symbol & Distribution \\
    \hline \\
    Mean max log viral load & $\mu_{v_\text{max}, g}$ & $\N(15, 15^2)$ \\
    Standard deviation of max log viral load & $\sigma_{v_\text{max}}$ & $\N_+(0, 10^2)$ \\
    Mean rate of log viral load increase & $\mu_{a, g}$ & $\N(1.25, 0.75^2)$ \\
    Standard deviation of rate of log viral load increase & $\sigma_{a}$ & $\N_+(0, 1)$ \\
    Mean rate of log viral load decrease & $\mu_{b, g}$ & $\N(0.5, 1.4^2)$ \\
    Standard deviation of rate of log viral load decrease & $\sigma_{b}$ & $\N_+(0, 1)$ \\
    Correlation of viral load parameters & $\matr{C}_\phi$ & $\LKJ(1)$ \\
    Negative log of rate of test failure & $-\log \rho$ & $\N(3, 0.5)$ \\
    Mean log viral load for test failures & $x_0$ & $\N(0, 3^2)$ \\
    Standard deviation of log viral load for test failures & $\sigma_0$ & $\N_+(3, 3^2)$ \\
    Standard deviation of log viral load observations & $\sigma_v$ & $1 + \N_+(2, 3^2)$ \\
    Time from first test to peak & $\tau_{\max,i}$ & $\N(5,5)$
\end{tabular}
\caption[Viral load model priors]{Priors for each parameter, taken from \textcite{hakkiOnset}. $\N_+$ indicates a normal distribution truncated to the non-negative reals. Details of the distributions, including their parameterisations, are in \cref{E-distributions}. The elements of $\vec{\mu}_{\phi,g}$ and $\vec{\sigma}_\phi$ have independent priors placed on them, as specified in the first lines of the table. Priors are the same for all $g$ and $i$. \label{ATACCC:table:hakki-priors}}
\end{table}
